Turbo fast no compromise gravitational wave parameter estimation

### 2022-12-12 src/tex/ms.tex
update heterodyned likelihood

\begin{align}
    \log{\mathcal{L}(d|\theta)} = -\left<d-h(\mathbf{\theta})|d-h(\mathbf{\theta})\right>/2,
\label{eq: loglikelihood}
\end{align}
where $d$ is the strain data, $h(\mathbf{\theta})$ is the strain predicted by a
waveform model with a specific set of source parameters. The right handside of
eq. \ref{eq: loglikelihood} can be evaluated in either time or frequency domain.
Since it is computationally cheaper to compute the likelihood in the frequency
domain, we choose to compute the likleihood throughout this work, which can be defined as 
\begin{align}
    \left<a|b\right> = 4 Re\int \frac{a^*(f)b(f)}{\mathcal{S}_n(f)} df,
\label{eq: innerproduct}
\end{align}
where $\mathcal{S}_n(f)$ is the noise power spectral density (PSD). To compute
the integral shown in eq. \ref{eq: innerproduct}, we need to evaluate a chosen
waveform model $h(\mathbf{\theta})$ at a number of frequnecy sample points.
Evaluting the likelihood function is often the most computationally intensive
part of the PE. The most accurate waveform model is numerical relativity (NR).
However, depending on the source parameters, generating one time series of
strain using can takes a day to half a year, which makes NR prohibitably
expensive for PE. To circumvent this problem, there are a number of waveform
"approximant" families, including the IMRPhenom family \kw{}, SEOB family \kw{},
and NR surrogate family\kw{cite}. For shorter event such as a $30-30\ M_{\odot}$
BBH, one waveform call could take $10\text{ms}$ to $\sim 1s$ \kw{verify number
in lalsuite}. For longer event such as a $1.4-1.4\ M_{\odot}$ BNS event, the
evaluation time could go up to \kw{Fill}. Since one needs to evaluate the
likelihood many times during sampling\footnote{A typical PE run with
\texttt{Bilby} takes $>10^6$ likelihood evalutation to converge.}, the
computational cost in evaluating the waveform accumulate and is the main reason
of the long runtime of GW PE.

@@

Since the computational cost of evaluating a waveform model scales linearly with
the number of sample points either in time or frequency domain, the
computational burden for longer-duration signals are often quite large. To
reduce the computational cost, there are a number of methods to reduce the
number of basis points one would need to compute the likelihood faithfully
\kw{cite}. We use heterodyned likelihood \kw{cite} (also named relative binning
in \kw{cite}) in this work. We give a concise description of what we implement
in our code here. For more extensive discussion of the derivation of heterodyned
likelihood, we refer the reader the reference \kw{}.

\kw{Specify the idea of heterodyned likelihood here}. The idea behind
heterodyned likelihood can be summarized as the following: the integrand in eq.
\ref{eq: innerproduct} is highly oscillatory function, so one has to sample the
integrand with sufficiently dense sampling to compute the integral faithfully.
The number of sample points needed would be much smaller if the integrand was
smooth. Given a pair of waveform parameters $\mathbf{\theta}$ and
$\mathbf{\theta_0}$ that are close to each other, the waveforms generated using
the pair of parameters are similar to each other, this means the ratio between
the waveforms is a smoothly varying function. Given a reference waveform
$h(\mathbf{\theta_0})$, we can exploit this similarity between waveforms to
reduce the number of sample points needed to compute the likelihood for the set
of $\mathbf{\theta}$ that are similar to $\mathbf{\theta_0}$. We decompose
the integrand into two parts: a highly oscillatory part that is depends only on
the reference waveform and the data, hence it only needs to be evaluated once;
and a smoothly varying part that depends on the target waveform parameters
$\theta$, that needs to be evaluated for every new likelihood evaluation.
Because the part that depends on the target waveform parameters is smooth, we
can use far fewer sample points to compute the integral with sufficient
accuracy.

\begin{align}
r(f) = \frac{h(f)}{h_0(f)} = r_0(h,b) + r_1(h,b)(f- f_m(b)) + \cdots
\end{align}

\begin{align}
    \langle d|h \rangle &\approx \sum_b A_0(b) r^*_0(h,b) + A_1(b) r^*_1(h,b) \nonumber \\
    \langle h|h \rangle &\approx \sum_b B_0(b) |r_0(h,b)|^2 + 2 B_1(b) \Re[r_0(h,b)r_1(h,b)]
\end{align}

@@

>>>>

\begin{align}
    \log{\mathcal{L}(d|\theta)} = -\left<d-h(\mathbf{\theta})|d-h(\mathbf{\theta})\right>/2,
\label{eq: loglikelihood}
\end{align}
where $d$ is the strain data, $h(\mathbf{\theta})$ is the strain predicted by a
waveform model with a specific set of source parameters. The right handside of
eq. \ref{eq: loglikelihood} can be evaluated in either time or frequency domain.
Since it is computationally cheaper to compute the likelihood in the frequency
domain, we choose to compute the likleihood throughout this work, which can be defined as 
\begin{align}
    % \left<a|b\right> = 4 Re\int \frac{a^*(f)b(f)}{\mathcal{S}_n(f)} df,
    \left<a|b\right> = 4 \int \frac{a^*(f)b(f)}{\mathcal{S}_n(f)} df,
\label{eq: innerproduct}
\end{align}
where $\mathcal{S}_n(f)$ is the noise power spectral density (PSD). To compute
the integral shown in eq. \ref{eq: innerproduct}, we need to evaluate a chosen
waveform model $h(\mathbf{\theta})$ at a number of frequnecy sample points.
Evaluting the likelihood function is often the most computationally intensive
part of the PE. The most accurate waveform model is numerical relativity (NR).
However, depending on the source parameters, generating one time series of
strain using can takes a day to half a year, which makes NR prohibitably
expensive for PE. To circumvent this problem, there are a number of waveform
"approximant" families, including the IMRPhenom family \kw{}, SEOB family \kw{},
and NR surrogate family\kw{cite}. For shorter event such as a $30-30\ M_{\odot}$
BBH, one waveform call could take $10\text{ms}$ to $\sim 1s$ \kw{verify number
in lalsuite}. For longer event such as a $1.4-1.4\ M_{\odot}$ BNS event, the
evaluation time could go up to \kw{Fill}. Since one needs to evaluate the
likelihood many times during sampling\footnote{A typical PE run with
\texttt{Bilby} takes $>10^6$ likelihood evalutation to converge.}, the
computational cost in evaluating the waveform accumulate and is the main reason
of the long runtime of GW PE.

@@

Since the computational cost of evaluating a waveform model scales linearly with
the number of sample points either in time or frequency domain, the
computational burden for longer-duration signals are often quite large. To
reduce the computational cost, there are a number of methods to reduce the
number of basis points one would need to compute the likelihood faithfully
\kw{cite}. We use heterodyned likelihood \kw{cite} (also named relative binning
in \kw{cite}) in this work. The idea behind heterodyned likelihood can be
summarized as the following: the integrand in eq. \ref{eq: innerproduct} is
highly oscillatory function, so one has to sample the integrand with
sufficiently dense sampling to compute the integral faithfully. The number of
sample points needed would be much smaller if the integrand was smooth. Given a
pair of waveform parameters $\mathbf{\theta}$ and $\mathbf{\theta_0}$ that are
close to each other, the waveforms generated using the pair of parameters are
similar to each other, this means the ratio between the waveforms is a smoothly
varying function. Given a reference waveform $h(\mathbf{\theta_0})$, we can
exploit this similarity between waveforms to reduce the number of sample points
needed to compute the likelihood for the set of $\mathbf{\theta}$ that are
similar to $\mathbf{\theta_0}$. We decompose the integrand into two parts: a
highly oscillatory part that is depends only on the reference waveform and the
data, hence it only needs to be evaluated once; and a smoothly varying part that
depends on the target waveform parameters $\theta$, that needs to be evaluated
for every new likelihood evaluation. Because the part that depends on the target
waveform parameters is smooth, we can use far fewer sample points to compute the
integral with sufficient accuracy.

@@

We give a concise description of what we implement
in our code here. For more extensive discussion of the derivation of heterodyned
likelihood, we refer the reader the reference \cite{Zackay:2018qdy}.

In the heterodyned likelihood framework, the two terms in eq \ref{eq:
loglikelihood} are given by
\begin{align}
    \langle d|h \rangle &\approx \sum_b A_0(b) r^*_0(h,b) + A_1(b) r^*_1(h,b), \nonumber \\
    \langle h|h \rangle &\approx \sum_b B_0(b) |r_0(h,b)|^2 + 2 B_1(b) \Re[r_0(h,b)r_1(h,b)],
\label{eq: heterodynedlikelihood}
\end{align}

where $b$ denotes the index of a \textit{sparse} set of bins where the integrand
will be computed on, $A_0(b)$, $A_1(b)$, $B_0(b)$, and $B_1(b)$ are the
coefficients of computed using the data and the reference waveform, and
$r_0(h,b)$ and $r_1(h,b)$ are the ratio between the target waveform and the
reference waveform at the center of the bin. To evaluate eq. \ref{eq:
heterodynedlikelihood}, we need to be able to first choose a binning scheme,
evaluate the coefficients given the data and the reference waveform, and finally
evaluate the ratio between the target waveform and the reference waveform at the
center of the bin.

Considering the phasing of a waveform is denoted by a power series $\Psi(f) =
\sum_i \alpha_i f^{\gamma_i}$, where $\alpha_i$ are some coefficients depending
on the waveform parameters and $\gamma_i$ are powers motivated by post-Newtonian
thoery. For example, for the term $\gamma_i = -5/3$, $\alpha_i$ is related to
the chirp mass. The maximum dephasing one can have within a frequency interval $[f_{\textrm{min}},f_{\textrm{max}}]$ is given by
\begin{align}
    \delta \Psi_{\textrm{max}}(f) = 2\pi \sum_{i} (f/f_{*,i})^{\gamma_i} \textrm{sgn}(\gamma_i),
\label{eq: maxdephasing}
\end{align}
where $f_{*,i} = f_{\textrm{max}}$ for $\gamma_i>=0$ and $f_{*,i} =
f_{\textrm{min}}$ for $\gamma_i<0$. Given the relation shown in eq. \ref{eq:
maxdephasing}, we can choose the binning scheme which devide the entire
frequency domain of interests into a set of bins such that the maximum dephasing
within each bin is smaller than a certain threshold $\epsilon$, i.e.
$|\delta\Psi_{\textrm{max}}(f_{\textrm{max}}) -
\delta\Psi_{\textrm{max}}(f_{\textrm{min}})| < \epsilon$.

Given a sufficiently dense binning scheme, the ratio between the target waveform
and the reference within a bin can be approximated as by linear interpolation
\begin{align}
r(f) = \frac{h(f)}{h_0(f)} = r_0(h,b) + r_1(h,b)(f- f_m(b)) + \cdots,
\label{eq:definer}
\end{align}
where $b$ is the index of a particular bin, $r_0(h,b)$ and $r_1(h,b)$ are the
value and slope of the ratio at the center of the bin resepctively, and
$f_m(b)$ is the center frequency of the bin. Since we have access to both $h(f)$
and $h_0(f)$, we can compute $r_0$ and $r_1$ by evaluating the value of $r(f)$
at the edge of the bin and invert eq. \ref{eq:definer}.

\begin{align}
    A_0(b) &= 4 \sum_{f \in b} \frac{d(f)h^*_0(f)}{S_n(f)} df, \\
    A_1(b) &= 4 \sum_{f \in b} \frac{d(f)h^*_0(f)(f-f_m(b))}{S_n(f)} df, \\
    B_0(b) &= 4 \sum_{f \in b} \frac{|h_0(f)|^2}{S_n(f)} df, \\
    B_1(b) &= 4 \sum_{f \in b} \frac{|h_0(f)|^2(f-f_m(b))}{S_n(f)} df,
\end{align}
